Summer 2025 ToDo

### 2025-10-10 Relations/Paper/Well-foundedness.tex
work on well-foundedness

An abstract reduction $\to_R$ is strongly normalizing
if and only if the converse relation is well-founded.
In this section, we will compare several definitions of well-foundedness,
which depend on the basic notions given in Definition~\ref{def:WFnotions} below.

@@

\begin{definition} \label{def:WFproperties} \hfill 
  \begin{enumerate}
    \item $R$ is \emph{accessibly well-founded} ($\WFacc$) if every element is accessible:
      \[
        \forall x \in A.\; x \text{ is accessible}
      \]
    \item $R$ is \emph{inductively well-founded} ($\WFind$) if every inductive predicate is universally true:
      \[
        \forall P \subseteq A.\;\text{$P$ is inductive} \Rightarrow \forall x \in A.\; x \in P
      \]
    \item $R$ is \emph{coreductively well-founded} ($\WFcor$) if every coreductive predicate is universally true:
      \[
        \forall P \subseteq A.\;\text{$P$ is coreductive} \Rightarrow \forall x \in A.\; x \in P
      \]
    \item  $R$ is \emph{well-founded minimality-wise} ($\WFmin$) if every nonempty subset contains a minimal element:
      \[
        \forall P \subseteq A.\; P \neq \varnothing \Rightarrow
        \exists x \in A.\; \text{$x$ is $P$-minimal}
      \]
    \item  $R$ is \emph{well-founded minimality-wise for $\lnot \lnot$-closed predicates} ($\WFminDNE$) if every nonempty $\lnot \lnot$-closed subset contains a minimal element:
      \[
        \forall P \subseteq A.\; P \neq \varnothing \Rightarrow \ol{\ol P} \subseteq P \Rightarrow
        \exists x \in A.\; \text{$x$ is $P$-minimal}
      \]
   \item $R$ is \emph{sequentially well-founded} ($\WFseq$) if every sequence contains an index at which it is not decreasing:
      \[
        \forall s : \nat \to A\ \exists k : \nat.\, \lnot \, R \,(s\,k)\,(s\,(k+1))
      \]
  \end{enumerate}
\end{definition}
We have formalized the above definitions of well-foundedness in \texttt{WFDefinitions.agda}.

@@

On the other hand, assuming accessibility is $\lnot\lnot$-closed, the other
hypothesis can now be weakened, to only require existence of minimal elements
for $\lnot\lnot$-closed subsets.  This is a significant restriction, which may
be validated in many cases of interests.  In contrast, the original hypothesis
(postulating minimal elements for ALL predicates)
is generally equivalent to every subset being decidable.
For example, the file \texttt{WFCounters.agda} shows that assuming the full minimality hypothesis
for the standard strict order on the natural numbers entails decidability
of every subset of $\nat$ --- which is clearly incompatible with effective semantics.

One could also consider proving well-foundedness ``up to $\lnot\lnot$'',
asserting that every element is $\lnot\lnot$-accessible,
every sequence $\lnot\lnot$-contains a non-decreasing index,
for every non-empty subset there $\lnot\lnot$-exists a minimal element, etc.
These are formalized in the definition below, and in the corresponding file
\texttt{WFWeakDefinitions.agda}. The logical relationships between all notions of well-foundedness are presented in Figure~\ref{fig:WF}.

@@

\begin{figure}[hb]
\centering  
\begin{tikzpicture}[
  scale=0.7,
  every node/.style={transform shape},
  node distance=1cm and .5cm,
  box/.style={draw, rectangle, rounded corners, minimum width=1cm, minimum height=1cm, align=center},
  arrow/.style={-{Latex}, thick}
]

@@

In \cite{Terese}, the $\WFseqm$ definition is used, as in the classical mathematics textbooks.
\apcomment{Move this comment elsewhere? Perhaps we should have a separate discussion between isWFseq- and isWFseq-not-not?}

The logical relationships between the various definitions of well-foundedness are summarized in 
Figure \ref{fig:WF}.
Again, all of the definitions appearing in this figure are classically equivalent.
But constructively, some implications are not provable in general.
Formalizing these implications enabled us to identify where exactly classical logic 
is used, and how much of it is truly needed.
An arrow from one node to another with no label indicates direct logical implication,
provable constructively with no additional assumptions. An arrow with a label indicates an additional 
(semi-)classical hypothesis needed to establish the implication. All of these facts have been 
formalized within the WellFounded subdirectory of the Agda repository.

{
\def\arraystretch{1.3}
\begin{figure}[h!]
\small 
\begin{tabular}{@{}l l l @{}}
\toprule
\textbf{Property} & \textbf{Definition}  &\textbf{Agda Code}  \\
\midrule
$\rdec$   & $R$ is decidable 
          & $\forall$ \verb|{x} {y}| $\to$ \verb|EM (R x y)|\\
$\FB$     & $R$ is finitely branching 
          & $\mathtt{\forall (a : A) \to \Sigma [ xs \in List A ] (\forall b \to R a b \to b \in{}List\, xs)}$ \\
$\accCor$ & Accessibility is coreductive 
          & \verb|R -coreductive (R -accessible)| \\
$\accDNE$ & Accessibility is $\lnot\lnot$-closed
          & $\lnot\lnot$\verb|Closed (R -accessible)| \\
$\corDNE$ & Coreductives are $\lnot\lnot$-closed 
          & $\mathtt{\forall (P : \mathcal{P} A) \to R}$-$\mathtt{coreductive\ P \to \lnot\lnot Closed\ P}$\\ 
$\mpe$    & Sequences are $\lnot\lnot$-closed
          & $\forall\ \mathtt{(f : \nat \to A) \to \lnot\lnot Closed 
          (\lambda x \to \Sigma [ k \in \nat ] (f\, k \equiv x))}$ \\
\bottomrule
\end{tabular}
\centering
\caption{Classical Properties Used in the Well-Foundedness Diagram}
\label{tab:cprop}
\end{figure}
}

@@

\apcomment{Things to do:}
\begin{itemize}
  \item Reiterate/articulate the role of classical principles in establishing relationships between definitions of 
    well-foundedness constructively 
  \item List/discuss? classical principles appearing in the figure 
  \item Discuss the figure overall -- refer to the enumerated list below 
  \item Create ``narrative bridges'' between these items 
  \item Final thing: Instantiate some of the classical principles on simple examples of term rewrite systems.
\end{itemize}

% As the figure illustrates, the concept of well-foundness is very rich from the vantage point of constructive logic.
% Our analysis resulted in the following principles, 
%
% The additional classical hypotheses needed to recover some of the implications on the diagram 
%
% The principles appearing in the labels of Figure \ref{fig:WF} are valid for many rewriting systems,
% and are classically true.  
%
% The weaker definitions help us to explore how the standard well-founded definitions interrelate. This is made clear in the following diagram.
%
\subsection{New version: Synopsis and Outline} 
This cuts down some, but not all, of the complexity.
For example, the two minimality principles
described above are equivalent with this formulation.
% The implication from the inductive to sequential definition no longer requires
% decidability of $R$.
% On the other hand, the reverse
At the same time, the
implication from sequential to inductive wellfoundedness
still requires an additional hypothesis, which may be described as the
constructive contrapositive of the definition of accessibility.
Specifically, we call {$\accCor$} the statement that, if an element $x$
is \emph{not} accessible, then there (strongly) exists a $y$ with $Ryx$
and $y$ not accessible.  (In the context of rewriting systems,
this condition asserts the existence of a function which maps any
non-strongly normalizing element to a one-step reduct with the same property.)
Assuming $\accCor$, all notions of well-foundedness
are constructively equivalent up to $\lnot\lnot$.

As this discussion illustrates, the network of relationships between the
constructive notions of well-foundedness is quite rich.  The purpose of
this section is to present the basic picture of these implications,
and elaborate on some of them.  Our key observations are summarized
in the following list.
\begin{enumerate}
  \item The implication from the inductive to the sequential definition
  requires the underlying relation to be decidable. This assumption is no longer
  necessary on the $\lnot\lnot$-translated side.
  \item WFmin is a very strong assumption and implies WFseq unconditionally.
  The more reasonable WFminDNE requires Markov's Principle
  (and decidability of $R$) to yield the same consequence.
  \item On the $\lnot\lnot$-side, WFmin- and WFminDNE- are equivalent,
  and imply WFacc- via WFseq-, requiring {$\accCor$} assumption in the final inference.
  \item At the same time, if the relation is finitely branching, then WFminDNE-
  implies WFacc- directly.
  \item Finite branchingness together with (weak) decidability of minimality
  implies $\accCor$.
  % \apcomment{The previous observation suggests the question of whether isWFseq- might
  % imply isWFminDNE- via decidability of minimal elements.\\
  % However, in trying to prove this implication, the hypothesis asserting non-existence of
  %  a minimal element of some given non-empty $\lnot\lnot$-closed predicate
  %  automatically entails that any particular element of interest is not normal.\\
  %  Unless it's not satisfying $P$? \emph{We need to decide whether it is P-minimal as well.}}
  \item Assuming $\accCor$, $\isWFacc$ ($\isWFaccm$) can also be derived from a weaker condition, $\isWFmin$ ($\isWFminm$).
  (Just like $\isWFminDNE$ is a natural weakening of $\isWFmin$ that is sufficient to recover $\isWFacc$ in the presence of {$\accDNE$}.)
  It is implied by $\isWFseqm$, but does not appear to imply anything on its own.

  % \apcomment{(This seems easy: The converse implication isWFminCor to isWFseq- requires the sequence in question to
  % have decidable image, leading to a further weakening of isWFseq-.)}
  \item Since isWFminDNE- implies isWFseq-, the former also implies isWFminCor-.
  But the converse fails: If a subset/predicate $P$ is $\lnot\lnot$-closed,
  we cannot conclude it is coreductive.

@@

  \item Assuming normal forms are decidable,
  isWFseq implies a very weak version of minimality, isWFminEM.
  On its own this property is insufficient to imply anything else,
  so we have nothing else to say about it.
\end{enumerate}

@@

>>>>

An abstract reduction $\to_R$ is strongly normalizing
if and only if the converse relation is well-founded.
In this section, we will compare several definitions of well-foundedness.
%which depend on the basic notions given in Definition~\ref{def:WFnotions} below.

\subsection{Definitions of well-foundedness}

@@

\begin{definition} \label{def:WFproperties} \hfill 
  \begin{enumerate}
    \item $R$ is \emph{accessibly well-founded} ($\WFacc$) if every element is accessible:
      \[
        \forall x \in A.\; x \text{ is accessible}
      \]
    \item $R$ is \emph{inductively well-founded} ($\WFind$) if every inductive predicate is universally true:
      \[
        \forall P \subseteq A.\;\text{$P$ is inductive} \Rightarrow \forall x \in A.\; x \in P
      \]
    \item $R$ is \emph{coreductively well-founded} ($\WFcor$) if every coreductive predicate is universally true:
      \[
        \forall P \subseteq A.\;\text{$P$ is coreductive} \Rightarrow \forall x \in A.\; x \in P
      \]
    \item  $R$ is \emph{well-founded minimality-wise} ($\WFmin$) if every nonempty subset contains a minimal element:
      \[
        \forall P \subseteq A.\; P \neq \varnothing \Rightarrow
        \exists x \in A.\; \text{$x$ is $P$-minimal}
      \]
    \item  $R$ is \emph{well-founded minimality-wise for $\lnot \lnot$-closed predicates} ($\WFminDNE$) if every nonempty $\lnot \lnot$-closed subset contains a minimal element:
      \[
        \forall P \subseteq A.\; P \neq \varnothing \Rightarrow \ol{\ol P} \subseteq P \Rightarrow
        \exists x \in A.\; \text{$x$ is $P$-minimal}
      \]
   \item $R$ is \emph{sequentially well-founded} ($\WFseq$) if every sequence contains an index at which it is not decreasing:
      \[
        \forall s : \nat \to A\ \exists k : \nat.\, \lnot \, R \,(s\,(k+1))\,(s\,k)
      \]
  \end{enumerate}
\end{definition}
We have formalized the above definitions of well-foundedness in \texttt{WFDefinitions.agda}.

@@

On the other hand, assuming accessibility is $\lnot\lnot$-closed, the other
hypothesis can now be weakened, to only require existence of minimal elements
for $\lnot\lnot$-closed subsets.  This is a significant restriction, which may
be validated in certain cases.  In contrast, the original hypothesis
(postulating minimal elements for ALL predicates)
is generally equivalent to every subset being decidable.
Indeed, the file \texttt{WFCounters.agda} shows that assuming the full minimality hypothesis
for the standard strict order on the natural numbers entails decidability
of every subset of $\nat$ --- which is clearly incompatible with effective semantics.
The restriction to $\lnot\lnot$-closed subsets implies (and is equivalent to) 
every subset $P$ being merely \emph{weakly} decidable (satisfying $\lnot \lnot P(x) \lor \lnot P(X)$ 
for all $x$).

One could also consider proving well-foundedness ``up to $\lnot\lnot$'',
asserting that every element is $\lnot\lnot$-accessible,
every sequence $\lnot\lnot$-contains a non-decreasing index,
for every non-empty subset there $\lnot\lnot$-exists a minimal element, etc.
These are formalized in the definition below, and in the corresponding file
\texttt{WFWeakDefinitions.agda}.

@@

In \cite{Terese}, the $\WFseqm$ definition is used, as in the classical mathematics textbooks.
\apcomment{Perhaps we should have a separate discussion between isWFseq- and isWFseq-not-not?}

\subsection{Implications between the definitions}

% The logical relationships between all of the notions of well-foundedness defined 
% above are presented in Figure~\ref{fig:WF}.
%
\begin{figure}[hb]
\centering  
\begin{tikzpicture}[
  scale=0.7,
  every node/.style={transform shape},
  node distance=1cm and .5cm,
  box/.style={draw, rectangle, rounded corners, minimum width=1cm, minimum height=1cm, align=center},
  arrow/.style={-{Latex}, thick}
]

@@

All of the above definitions of well-foundedness are classically equivalent.
But constructively, the only provable equivalence is between the inductive and accessible 
formluations (reflecting the fact that accessibility is defined as the least inductive predicate).
Formalizing the remaining implications enabled us to identify where exactly classical logic 
is used, and how much of it is truly needed.

The logical relationships between the various definitions are summarized in Figure \ref{fig:WF}.
An arrow from one node to another with no label indicates direct logical implication,
provable constructively with no additional assumptions. An arrow with a label indicates an additional 
(semi-)classical hypothesis needed to establish the implication. 
The definitions of these hypotheses are given in Figure \ref{tab:cprop}.
All of the implications have been 
formalized within the WellFounded subdirectory of the Agda repository.

@@

% As the figure illustrates, the concept of well-foundness is very rich from the vantage point of constructive logic.
% Our analysis resulted in the following principles, 
%
% The additional classical hypotheses needed to recover some of the implications on the diagram 
%
% The principles appearing in the labels of Figure \ref{fig:WF} are valid for many rewriting systems,
% and are classically true.  
%
% The weaker definitions help us to explore how the standard well-founded definitions interrelate. This is made clear in the following diagram.
%
As can be seen from the figure, taking the $\lnot\lnot$-closures of 
well-foundedness properties cuts down some, but not all, of the complexity.
For example, the two minimality principles
described above are equivalent with this formulation.
% The implication from the inductive to sequential definition no longer requires
% decidability of $R$.
% On the other hand, the reverse
At the same time, the weak version of the sequential definition is unable to imply 
the (weak version of) accessible one without an additional hypothesis $\accCor$, 
which may be described as the constructive contrapositive of the definition of accessibility.
In the context of rewriting systems,
this condition asserts the existence of a function which maps any
non-strongly normalizing element to a one-step reduct with the same property.
Assuming $\accCor$, all notions of well-foundedness
are constructively equivalent up to $\lnot\lnot$.

As the Figure \ref{tab:cprop} illustrates, the network of relationships between the
constructive notions of well-foundedness is quite rich. 
The main observations can now be read off from the figure.
\begin{enumerate}
  \item The implication from $\isWFacc$ to $\isWFseq$ requires the underlying relation to be decidable. 
    This assumption is no longer necessary on the $\lnot\lnot$-translated side.
  \item WFmin is a very strong assumption and implies WFseq unconditionally.
  The more reasonable WFminDNE requires the image of every sequence to be $\lnot\lnot$-closed 
  to complete the implication.  This condition can be considered as a special case of Markov's Principle
  (assuming decidability of equality on $A$).
  \item On the $\lnot\lnot$-side, WFmin- and WFminDNE- are equivalent,
  and imply WFacc- via WFseq-, requiring {$\accCor$} assumption in the final inference.
  \item At the same time, if the relation is finitely branching ($\FB$), then WFminDNE-
  implies WFacc- directly.  It should be noted that, $\FB$ together with (weak) decidability of minimality
  implies $\accCor$.  \apcomment{Fix this: This is proved in \texttt{FinitelyBranching.agda}.}.
  % \apcomment{The previous observation suggests the question of whether isWFseq- might
  % imply isWFminDNE- via decidability of minimal elements.\\
  % However, in trying to prove this implication, the hypothesis asserting non-existence of
  %  a minimal element of some given non-empty $\lnot\lnot$-closed predicate
  %  automatically entails that any particular element of interest is not normal.\\
  %  Unless it's not satisfying $P$? \emph{We need to decide whether it is P-minimal as well.}}
  \item Assuming $\accCor$, $\isWFacc$ ($\isWFaccm$) can also be derived from a weaker condition, $\WFcor$ ($\WFcorm$).
    \apcomment{Add the ``is''!}
    (Just like $\isWFminDNE$ is a natural weakening of $\isWFmin$ that is sufficient to recover $\isWFacc$ in the presence of {$\accDNE$}.)
  \apcomment{True?  It is implied by $\isWFseqm$, but does not appear to imply anything on its own.}

  % \apcomment{(This seems easy: The converse implication isWFminCor to isWFseq- requires the sequence in question to
  % have decidable image, leading to a further weakening of isWFseq-.)}
\item \apcomment{Fix with macros}. Since isWFminDNE- implies isWFseq-, the former also implies isWFminCor-.
  But the converse fails: If a subset/predicate $P$ is $\lnot\lnot$-closed,
  we cannot conclude it is coreductive.

@@

  % \item Assuming normal forms are decidable,
  % isWFseq implies a very weak version of minimality, isWFminEM.
  % On its own this property is insufficient to imply anything else,
  % so we have nothing else to say about it.
\end{enumerate}

@@

% \apcomment{Things to do:}
% \begin{itemize}
%   \item Reiterate/articulate the role of classical principles in establishing relationships between definitions of 
%     well-foundedness constructively 
%   \item List/discuss? classical principles appearing in the figure 
%   \item Discuss the figure overall -- refer to the enumerated list below 
%   \item Create ``narrative bridges'' between these items 
%   \item Final thing: Instantiate some of the classical principles on simple examples of term rewrite systems.
% \end{itemize}
%
%
